---
title: "Time Series Final Project"
author: "Shubhangi Sharma, Akanksha Arun, Juhil Ahir, Tanvi Vijay"
date: "2023-12-01"
output:
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Initial data exploration

```{r}
library(forecast)
library(dplyr)
library(ggplot2)

avocados = read.csv("avocado.csv",header = TRUE)

avocados_c = filter(avocados, region == "California",type=="conventional")
avocados_o = filter(avocados, region == "California",type=="organic")

avocados_c = avocados_c %>% arrange(Date)
avocados_o = avocados_o %>% arrange(Date)

tdx = c(1:169)/52+2015
plot(tdx,avocados_c$Total.Volume/1000,type="l",ylab="Total Volume sold (K)",xlab="Time",main = "Conventional Avocados")
plot(tdx,avocados_o$Total.Volume/1000,type="l",ylab="Total Volume sold (K)",xlab="Time",main = "Organic Avocados")
#based on the initial plot we see no trend, seasonality, level and noise

avocados_c.ts = ts(avocados_c[,4], frequency = 52, start=c(2015))
avocados_o.ts = ts(avocados_o[,4], frequency = 52, start=c(2015))

autoplot(decompose(avocados_c.ts))
autoplot(avocados_c.ts/1000,ylab="Total Volume sold (K)",xlab="Time",main = "Conventional Avocados")
ts1 = aggregate(avocados_c.ts,FUN=mean)
autoplot(ts1)

# checking seasonality
Acf(avocados_c.ts,lag.max=150)
# We see a peak at lag 52 that implies a yearly seasonality component


autoplot(decompose(avocados_o.ts))
autoplot(avocados_o.ts/1000,ylab="Total Volume sold (K)",xlab="Time",main = "Organic Avocados")
ts2 = aggregate(avocados_o.ts,FUN=mean)
autoplot(ts2)

# checking seasonality
Acf(avocados_o.ts,lag.max=150)
# We see a yearly seasonality component


boxplot(avocados_c.ts)
#we have outliers
boxplot(avocados_c.ts/1000~cycle(avocados_c.ts),xlab = "Month",ylab="Sales")
# on average across all years we see a high volume of conventional avocados being sold in the second week of February
seasonplot(avocados_c.ts/1000,year.labels=T,col=c("blue","red","green","purple"),lwd=2)
```

```{r}
boxplot(avocados_o.ts)
boxplot(avocados_o.ts/1000~cycle(avocados_o.ts),xlab = "Month",ylab="Sales")
seasonplot(avocados_o.ts/1000,year.labels=T,col=c("blue","red","green","purple"),lwd=2)
# Similar to conventional, we could observe a increase in sale for organic avocado during the time frame of the 9th to 12th week.
```


# Creating training and testing data for Conventional Avocado
```{r}
n = length(avocados_c.ts)
n
# set up the forecasting horizon
stepsAhead = 12 

# assign the length of the training set
nTrain = n - stepsAhead

#create the training set
train.ts = window(avocados_c.ts ,frequency = 52, end=c(2018))


#create the validation set
valid.ts = window(avocados_c.ts, frequency = 52, start=c(2018))
```

# Creating training and testing data for Organic Avocado
```{r}

#create the training set
train.ts_org = window(avocados_o.ts ,frequency = 52, end=c(2018))


#create the validation set
valid.ts_org = window(avocados_o.ts, frequency = 52, start=c(2018))

```

# Modeling

#1. Linear Models for Modeling (Conventional)

```{r}
# fitting a linear model with trend and seasonality which is additive
model1 = tslm(train.ts~trend + season)
model1
autoplot(train.ts) +
  geom_line(aes(y=model1$fitted.values),col=2)
checkresiduals(model1)
accuracy(model1)

# fitting a linear model with trend and seasonality which is multiplicative
model2 = tslm(log(train.ts)~trend + season)
model2
autoplot(log(train.ts)) + geom_line(aes(y=model2$fitted.values),col=3)
checkresiduals(model2)
accuracy(model2)

```

#1.1. Linear Models for Modeling (Organic)
```{r}
model1_org = tslm(train.ts_org~trend + season)
model1_org
autoplot(train.ts_org) +
  geom_line(aes(y=model1_org$fitted.values),col=2)
checkresiduals(model1_org)
accuracy(model1_org)

model2_org = tslm(log(train.ts_org)~trend + season)
model2_org
autoplot(log(train.ts_org)) + geom_line(aes(y=model2_org$fitted.values),col=3)
checkresiduals(model2_org)
accuracy(model2_org)
```

#2 Naive and Seasonal Naive for Conventional Avocado

```{r}
# use function naive to get 12-step-ahead forecast
naive = naive(train.ts,h=12)
naive
# use function snaive to create a 1-season-ahead forecast
seasonal = snaive(train.ts, 12)
seasonal

accuracy(naive, valid.ts)
accuracy(seasonal, valid.ts)

#3. moving average model
library(zoo)
nValid = length(valid.ts)
ma.centered = ma(train.ts, order = 3)

autoplot(train.ts, ylab = "Demand", col="black", series = "Observed")+ 
  autolayer(ma.centered, series = "Moving Average", lwd = 1.5)+
  scale_x_continuous(breaks = seq(from = 2003, to = 2006,6))

#4. trailing moving average

ma.trailing = rollmean(train.ts, k = 3, align= "right")
# Find the last moving average in the training period
pred = tail(ma.trailing, 1)

# Use the last moving average as the prediction for each month in the validation period
ma.trailing.pred = ts(rep(pred, nValid), start = start(valid.ts),frequency = frequency(valid.ts))

# let's find the actual observed average over the validation period

mean.valid = ts(rep(mean(valid.ts), nValid), start = start(valid.ts), frequency = frequency(valid.ts))

autoplot(train.ts, ylab = "Demand", col="black", series = "Observed")+ 
  autolayer(ma.centered, series = "Moving Average", lwd = 1.5)+
   autolayer(ma.trailing.pred, series = "Predicted", lwd=1.5)+
  autolayer(mean.valid, series = "Average", lwd = 1.5)
  #scale_x_continuous(breaks = seq(from = 2003, to = 2006))

accuracy(ma.trailing, train.ts)
accuracy(ma.trailing.pred, valid.ts)


#5. Holt Winter's model (Since there is seasonality) -> we can NOT apply because frequency is greater than 24
```

#2.1 Naive and Seasonal Naive for Organic Avocado
```{r}
# use function naive to get 12-step-ahead forecast
naive_org = naive(train.ts_org,h=12)
naive_org
# use function snaive to create a 1-season-ahead forecast
seasonal_org = snaive(train.ts_org, 12)
seasonal_org

accuracy(naive_org, valid.ts_org)
accuracy(seasonal_org, valid.ts_org)

#3. moving average model
library(zoo)
nValid = length(valid.ts_org)
ma.centered_org = ma(train.ts_org, order = 3)

autoplot(train.ts_org, ylab = "Demand", col="black", series = "Observed")+ 
  autolayer(ma.centered_org, series = "Moving Average", lwd = 1.5)+
  scale_x_continuous(breaks = seq(from = 2003, to = 2006,6))

#4. trailing moving average

ma.trailing_org = rollmean(train.ts_org, k = 3, align= "right")
# Find the last moving average in the training period
pred_org = tail(ma.trailing_org, 1)

# Use the last moving average as the prediction for each month in the validation period
ma.trailing.pred_org = ts(rep(pred_org, nValid), start = start(valid.ts_org),frequency = frequency(valid.ts_org))

# let's find the actual observed average over the validation period

mean.valid_org = ts(rep(mean(valid.ts_org), nValid), start = start(valid.ts_org), frequency = frequency(valid.ts_org))

autoplot(train.ts_org, ylab = "Demand", col="black", series = "Observed")+ 
  autolayer(ma.centered_org, series = "Moving Average", lwd = 1.5)+
   autolayer(ma.trailing.pred_org, series = "Predicted", lwd=1.5)+
  autolayer(mean.valid_org, series = "Average", lwd = 1.5)
  #scale_x_continuous(breaks = seq(from = 2003, to = 2006))

accuracy(ma.trailing_org, train.ts_org)
accuracy(ma.trailing.pred_org, valid.ts_org)


#5. Holt Winter's model (Since there is seasonality) -> we can NOT apply because frequency is greater than 24

```


•	Moving average of raw series
	◦	Would not be suitable. The moving average method can’t account for seasonality and lags behind trend
	•	Moving average of deseasonalized series
	◦	Would also not be suitable. While moving average would work better on the deseasonalized series than the raw series, it is still not ideal.
	•	Simple exponential smoothing of the raw series
	◦	Would not be suitable. As with the moving average method, simple exponential smoothing does not work well for series with seasonality and/or trend.
	•	Double exponential smoothing of the raw series
	◦	Would not be suitable. While double exponential smoothing could deal with the trend in the raw series it could not account for the seasonality.

#3. Checking for the auto correlation factors in the data

```{r}

par(mfrow=c(1,2))
Acf(avocados_c.ts, 105, main="Conventional Avocado") # tailing off
Pacf(avocados_c.ts, 105, main="") # mostly cuts off after 1 spike or tails off
par(mfrow=c(1,1))
```

```{r}
# For organic 
par(mfrow=c(1,2))
Acf(avocados_o.ts, 105, main="Organic Avocado") # tailing off
Pacf(avocados_o.ts, 105, main="") # mostly cuts off after 1 spike 
par(mfrow=c(1,1))
```

#4. ARIMA models for Conventional Avocados

```{r}
library(forecast)

#so we will try AR(1)
m1 = Arima(train.ts, order = c(1,0,0))
summary(m1)

#no need to difference because no trend 
#but seasonslity is there
m2 = Arima(train.ts, order = c(1,0,0), seasonal = list(order = c(1,0,0), period = 52))
summary(m2)
#prediction
m2.p = forecast(m2, h = length(valid.ts))
autoplot(m2.p)

autoplot(valid.ts)+ autolayer(m2.p$mean)

accuracy(m2.p, valid.ts)

checkresiduals(m2)

#model 3
m3 = Arima(train.ts, order = c(1,0,0), seasonal = list(order = c(0,1,0), period = 52))
summary(m3)


#prediction
m3.p = forecast(m3, h = length(valid.ts))
autoplot(m3.p)

autoplot(valid.ts)+ autolayer(m3.p$mean)

accuracy(m3.p, valid.ts)

checkresiduals(m3)

#model 4
m4 = Arima(train.ts, order = c(1,0,0), seasonal = list(order = c(1,1,0), period = 52))
summary(m4)


#prediction
m4.p = forecast(m4, h = length(valid.ts))
autoplot(m4.p)

autoplot(valid.ts)+ autolayer(m4.p$mean)

accuracy(m4.p, valid.ts)

checkresiduals(m4)
```

```{r}
# For Organic
#so we will try AR(1)
m1_org = Arima(train.ts_org, order = c(1,0,0))
summary(m1_org)

#no need to difference because no trend 
#but seasonslity is there
m2_org = Arima(train.ts_org, order = c(1,0,0), seasonal = list(order = c(1,0,0), period = 52))
summary(m2_org)
#prediction
m2.p_org = forecast(m2_org, h = length(valid.ts_org))
autoplot(m2.p_org)

autoplot(valid.ts_org)+ autolayer(m2.p_org$mean)

accuracy(m2.p_org, valid.ts_org) # MAPE = 16.34

checkresiduals(m2_org)

#model 3
m3_org = Arima(train.ts_org, order = c(1,0,0), seasonal = list(order = c(0,1,0), period = 52))
summary(m3_org)


#prediction
m3.p_org = forecast(m3_org, h = length(valid.ts_org))
autoplot(m3.p_org)

autoplot(valid.ts_org)+ autolayer(m3.p_org$mean)

accuracy(m3.p_org, valid.ts_org) #MAPE = 12.74

checkresiduals(m3_org)

#model 4
m4_org = Arima(train.ts_org, order = c(1,0,0), seasonal = list(order = c(1,1,0), period = 52))
summary(m4_org)


#prediction
m4.p_org = forecast(m4_org, h = length(valid.ts_org))
autoplot(m4.p_org)

autoplot(valid.ts_org)+ autolayer(m4.p_org$mean)

accuracy(m4.p_org, valid.ts_org) # MAPE = 11.15

checkresiduals(m4_org)
```


#5. Adding the effect of price to the model

```{r}
avocados_c_pr.ts = ts(avocados_c[,3], frequency = 52, start=c(2015))
autoplot(avocados_c_pr.ts, main="Price", ylab="", xlab="Years")


pairs(cbind(Volume = avocados_c.ts, Price = avocados_c_pr.ts))

price.tr = window(avocados_c_pr.ts ,frequency = 52, end=c(2018,1))
price.v = window(avocados_c_pr.ts ,frequency = 52, start=c(2018,2))

Ccf(avocados_c.ts, avocados_c_pr.ts, 12)


m1 = tslm(train.ts~trend) # weekly drop of 0.03433
m1.p = forecast(m1,h=12)
summary(m1)
accuracy(m1.p,valid.ts)
checkresiduals(m1.p)

#m2 = tslm(train.ts~trend + price.tr)
#m2.p = forecast(m2,h=12)
#summary(m2)
#accuracy(m2.p,valid.ts)
#checkresiduals(m2.p)



data.tr<-ts.intersect(train.ts, 
                      AveragePrice= price.tr)

m2 = tslm(train.ts~trend + AveragePrice, data=data.tr) 

summary(m2)

autoplot(train.ts)+
autolayer(m2$fitted.values)

#training error
accuracy(m2$fitted.values, train.ts)


data.valid=as.data.frame(ts.intersect(valid = valid.ts,
                      AveragePrice = price.v))
                      
pred=forecast(m2, newdata= data.valid, h=dim(data.valid)[1], level=0)

autoplot(valid.ts)+
  autolayer(pred)

accuracy(pred, valid.ts)
```

```{r}

# For Organic 

avocados_o_pr.ts = ts(avocados_o[,3], frequency = 52, start=c(2015))
autoplot(avocados_o_pr.ts, main="Price", ylab="", xlab="Years")


pairs(cbind(Volume = avocados_o.ts, Price = avocados_o_pr.ts))

price.tr_org = window(avocados_o_pr.ts ,frequency = 52, end=c(2018,1))
price.v_org = window(avocados_o_pr.ts ,frequency = 52, start=c(2018,2))

Ccf(avocados_o.ts, avocados_o_pr.ts, 12)


m1_org = tslm(train.ts_org~trend) 
m1.p_org = forecast(m1_org,h=12)
summary(m1_org)
accuracy(m1.p_org,valid.ts_org) # MAPE of 12.75
checkresiduals(m1.p_org)


data.tr_org<-ts.intersect(train.ts_org, 
                      AveragePrice= price.tr_org)

m2_org = tslm(train.ts_org~trend + AveragePrice, data=data.tr_org) 

summary(m2_org)

autoplot(train.ts_org)+
autolayer(m2_org$fitted.values)

#training error
accuracy(m2_org$fitted.values, train.ts_org)

data.valid_org=as.data.frame(ts.intersect(valid = valid.ts_org,
                      AveragePrice = price.v_org))
                      
pred_org=forecast(m2_org, newdata= data.valid_org, h=dim(data.valid_org)[1], level=0)

autoplot(valid.ts_org)+
  autolayer(pred_org)

accuracy(pred_org, valid.ts_org) # MAPE of 5.82
```


#5. Putting the best model values of the training data in the forecasting model with price

```{r}

train.ts = m4$fitted

fdata.tr<-ts.intersect(train.ts, 
                      AveragePrice= price.tr)

m2 = tslm(train.ts~trend + AveragePrice, data=fdata.tr) 

summary(m2)

autoplot(train.ts)+
  autolayer(m2$fitted.values)

#training error
accuracy(m2$fitted.values, train.ts)

data.valid=as.data.frame(ts.intersect(valid = valid.ts,
                      AveragePrice = price.v))
                      
pred=forecast(m2, newdata= data.valid, h=dim(data.valid)[1], level=0)

autoplot(valid.ts)+
  autolayer(pred)

accuracy(pred, valid.ts) # MAPE of 6.89

checkresiduals(pred)
```

```{r}
# For Organic
train.ts_org= m4_org$fitted

fdata.tr_org<-ts.intersect(train.ts_org, 
                      AveragePrice= price.tr_org)

m2_org = tslm(train.ts_org~trend + AveragePrice, data=fdata.tr_org) 

summary(m2_org)

autoplot(train.ts_org)+
autolayer(m2_org$fitted.values)

#training error
accuracy(m2_org$fitted.values, train.ts_org)

data.valid_org=as.data.frame(ts.intersect(valid = valid.ts_org,
                      AveragePrice = price.v_org))
                      
pred_org=forecast(m2_org, newdata= data.valid_org, h=dim(data.valid_org)[1], level=0)

autoplot(valid.ts_org)+
  autolayer(pred_org)

accuracy(pred_org, valid.ts_org)  # MAPE = 5.94

```

Based on the models we tried we can see that the model with ARIMA that we have plottted added with the effect of price is working the best on the validation set.
Further we will now use the complete dataset to train the model and forecast for the rest of 2018.

# For Conventional Avocados
```{r}
library(forecast)
final_train.ts = window(avocados_c.ts ,frequency = 52, start=c(2015))

final_arima = Arima(final_train.ts, order = c(1,0,0), seasonal = list(order = c(1,1,0), period = 52))
summary(final_arima)


#prediction
final_arima.p = forecast(final_arima, h = 52)
autoplot(final_arima.p)

autoplot(final_train.ts,main="Conventional Avocados Sales Forecast",ylab="Volume (K)")+ autolayer(final_arima.p$mean)

final_arima.p

min(final_arima.p$mean)
max(final_arima.p$mean)

accuracy(final_train.ts,final_arima.p$fitted)
```


# For Organic Avocados
```{r}
final_train.ts = window(avocados_o.ts ,frequency = 52, start=c(2015))

final_arima = Arima(final_train.ts, order = c(1,0,0), seasonal = list(order = c(1,1,0), period = 52))
summary(final_arima)


min(final_arima.p$mean)
max(final_arima.p$mean)
#prediction
final_arima.p = forecast(final_arima, h = 52)
autoplot(final_arima.p)

autoplot(final_train.ts,main="Organic Avocados Sales Forecast",ylab="Volume (K)")+ autolayer(final_arima.p$mean)

final_arima.p

accuracy(final_train.ts,final_arima.p$fitted)

```


